**{shorts}**: An R Package for Modeling Short Sprints

### 2022-07-07 shorts-paper.Rmd
Rewrote to be used with {shorts} 2.0.0

---
title: "shorts: An R Package for Modeling Short Sprints"
date: "`r Sys.Date()`"
author: 
  - Mladen Jovanović^[Faculty of Sport and Physical Education, University of Belgrade, Serbia, coach.mladen.jovanovic@gmail.com]
  - Jason D. Vescovi^[Faculty of Kinesiology and Physical Education, Graduate School of Exercise Science, Toronto, ON Canada, vescovij@gmail.com]
abstract: >
  Short sprint performance is one of the most distinguishable and admired physical traits in sports. Short sprints have been modeled using the mono-exponential equation that involves two parameters: (1) maximum sprinting speed (MSS) and (2) relative acceleration (TAU). The most common methods to assess short sprint performance are with a radar gun or timing gates. In this paper, we: 1) provide the **shorts** package that can model sprint timing data from these two sources; 2) discuss potential issues with assessing sprint time (synchronization and flying start, respectively); and 3) provide model definitions within the **shorts** package to help alleviate errors within the subsequent parameter outcomes.
citation_package: natbib
bibliography: [references.bib, packages.bib]
output:
  bookdown::word_document2:
    number_sections: FALSE
link-citations: yes
editor_options:
  chunk_output_type: console
---

@@

def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\n \\", options$size,"\n\n", x, "\n\n \\normalsize"), x)
})

@@

# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), "bookdown", "knitr", "rmarkdown", "shorts", "nlme", "LambertW", "kableExtra"
 ), "packages.bib")

@@

options(
  digits = 3,
  "width" = 65,
  scipen = 999
)
```

@@

Short sprint performance is one of the most distinguishable and admired physical traits in sports. Short sprints, commonly performed in most team sports (e.g., soccer, field hockey, handball, football, etc.), are defined as maximal running from a stand still position over a distance that doesn't result in deceleration at the end. Peak anaerobic power is achieved within the first few seconds (<5 s) of maximal efforts [@mangineSpeedForcePower2014], whereas the ability to achieve maximal sprint speed varies based on the athlete and their sport. For example, track and field sprinters are trained to achieve maximal speed later in a race (i.e., 50-60 m) [@ward-smithEnergyConversionStrategies2001], but team sport athletes have sport-specific attributes and reach it much sooner (i.e., 30-40 m)[@brownAssessmentLinearSprinting2004]. Regardless of the differences in kinematics between athletes, evaluating short sprint performance is routinely included within a battery of fitness tests for a wide range of sports

The use of force plates is considered the gold standard for assessing mechanical properties of sprinting; however, there are logistical and financial challenges to capturing the profile of an entire sprint [@morinSimpleMethodComputing2019; @samozinoSimpleMethodMeasuring2016]. Radar and laser technology are frequently used laboratory-grade methods [@buchheitMechanicalDeterminantsAcceleration2014; @edwardsSprintAccelerationCharacteristics2020; @jimenez-reyesRelationshipVerticalHorizontal2018; @marcote-pequenoAssociationForceVelocity2019] but not normally accessible to practitioners working in sports. Undoubtedly, the most common method available and used to evaluate sprint performance are timing gates. Often multiple gates are positioned at varying distances to capture split times (e.g., 5, 10, 20 m), which can now be incorporated into the method for determining sprint mechanical properties [@morinSimpleMethodComputing2019; @samozinoSimpleMethodMeasuring2016]. This approach presents an advantage to practitioners who can use the outcomes to describe individual differences, quantify the effects of training interventions, and better understanding the limiting factors of performance.

The **shorts** package [@R-shorts], written in the R language [@R-base], represents an open-source tool to help sports scientists translate raw timing data into detailed mechanical outcomes through mathematical modeling [@morinSimpleMethodComputing2019; @samozinoSimpleMethodMeasuring2016]. To best of our knowledge, scientist, researchers, and coaches have been performing short sprints modeling using the built-in solver function of Excel (Microsoft Corporation, Redmond, Washington, United States) [@morinSpreadsheetSprintAcceleration2017; @morinSpreadsheetSprintAcceleration2019; @stenrothForcevelocityProfilingIce2020; @stenrothSpreadsheetSprintAcceleration2020; @samozinoSimpleMethodMeasuring2016; @clarkNFLCombine40Yard2017; @morinSimpleMethodComputing2019], which makes the **shorts** package a major improvement in ease-of-use, speed, transparency, reproducibility, and more feature-rich model fitting.

In the current paper, we will provide an explanation of one commonly used mathematical equation to model short sprints, modeling applications using the **shorts** package, issues that can arise during measurement and estimation, and potential solutions to those problems.

@@

Short sprints have been modeled using the mono-exponential equation \@ref(eq:velocity-time) originally proposed by @doi:10.1098/rspb.1927.0035, and more recently popularized by @clarkNFLCombine40Yard2017, and @samozinoSimpleMethodMeasuring2016. Equation \@ref(eq:velocity-time) represents the function for instantaneous horizontal velocity $v$ given the time $t$ and two model parameters:

@@

The parameters of the equation \@ref(eq:velocity-time) are *maximum sprinting speed* (MSS; expressed in $ms^{-1}$) and *relative acceleration* (TAU). Mathematically, TAU represents the ratio of MSS to initial acceleration (MAC; *maximal acceleration*, expressed in $ms^{-2}$) \@ref(eq:maximal-acceleration).

@@

Although TAU is used in the equations, and later estimated, we prefer to use MAC instead since it is easier to grasp, particularly for less math inclined coaches.

@@

Let's consider four athletes with different levels of MSS (high versus low maximal sprinting speed) and MAC (high versus low maximal acceleration; as mentioned previously, using MAC is preferred over using TAU) (Table \@ref(tab:four-athletes-table)).

@@

knitr::kable(
  athletes,
  caption = "Four athletes with different MSS and MAC parameters.",
  digits = 2,
  booktabs = TRUE
)
```

Figure \@ref(fig:four-athletes-kinematics) depicts distance, velocity, and acceleration over time (from 0 to 6 s).

```{r four-athletes-kinematics, echo=FALSE, fig.cap="Kinematic characteristic of four athletes with different MSS and MAC parameters over a period of 0 to 6 seconds."}
kinematics <- expand_grid(
  athletes,
  time = seq(0, 6, length.out = 1000)
) %>%
  mutate(
    distance = predict_distance_at_time(time, MSS, TAU),
    velocity = predict_velocity_at_time(time, MSS, TAU),
    acceleration = predict_acceleration_at_time(time, MSS, TAU)
  )

@@

Plotting acceleration against velocity (Figure \@ref(fig:four-athletes-profile)), we will get *Acceleration-Velocity Profile*, which is linear, according to the mathematical model. If the athlete's body mass (kg) is known, as well as additional air resistance parameters (see [Air resistance and the calculation of force and mechanical power] section of this paper), *Force-Velocity Profile* can be estimated (see [Force-Velocity profile] section of this paper).

```{r four-athletes-profile, echo=FALSE, fig.cap="Acceleration-Velocity profile of four athletes with different MSS and MAC parameters."}
gg <- ggplot(
  kinematics,
  aes(x = velocity, y = acceleration, color = Athlete)
) +
  theme_bw(8) +
  geom_line(alpha = 0.7) +
  ylab(expression("Acceleration (ms"^-2 * ")")) +
  xlab(expression("Velocity (ms"^-1 * ")")) +
  xlim(c(0, 12)) +
  theme(
    legend.position = "top",
    legend.title = element_blank()
  )

@@

# Estimation using shorts package

Short sprints profiling is usually performed by: (1) measuring split times using timing gates (i.e., positioned at various distances, e.g., 5, 10, 20, 30, 40 m), or (2) getting a velocity trace using a radar gun. Estimation of MSS and TAU parameters from equation \@ref(eq:velocity-time) is performed in **shorts** package using non-linear least squares regression implemented in the `nls()` function [@batesNonlinearModels1992; @batesNonlinearRegressionAnalysis2007] in the **base R** [@R-base] and `nlme()` function in the **nlme** package [@R-nlme] for the mixed-effect models.

@@

Let's consider an example of an athlete with MSS equal to 9 $ms^{-1}$, TAU equal to 1.3, and MAC equal to 6.92 $ms^{-2}$ performing 40m sprint with timing gates positioned at each 10m split. For split times, distance is a predictor, and time is the outcome variable, thus the equation \@ref(eq:distance-time) becomes:

@@

$W$ in equation \@ref(eq:time-distance) represents Lambert's W function [@R-LambertW]. Researchers often incorrectly use the equation \@ref(eq:distance-time) [@morinSpreadsheetSprintAcceleration2017; @morinSpreadsheetSprintAcceleration2019; @stenrothSpreadsheetSprintAcceleration2020], in which the time is the predictor and distance is the outcome variable, instead of statistically correct equation \@ref(eq:time-distance). This practice should be avoided because switching of the predictor and outcome variables in the regression model might produce biased estimated parameters [@motulskyIntuitiveBiostatisticsNonmathematical2018, p. 341]. This might not necessarily be the case with the short sprints profiling, but it is nevertheless a bad statistical practice.

MSS and TAU parameters are estimated using `model_using_splits()` function:

```{r}
require(shorts)

split_distance <- c(10, 20, 30, 40)

split_time <- c(2.17, 3.43, 4.60, 5.73)

m1 <- model_using_splits(
  distance = split_distance,
  time = split_time
)

@@

Maximal relative power (PMAX) from the output is estimated using $\frac{MSS \times MAC}{4}$, which disregards the air resistance. `time_correction` and `distance_corection` parameters will be covered later in the paper.

Besides providing *residual standard error* (RSE), **shorts** functions provide additional model fit estimators. Additional information can be gained by exploring the returned object, particularly object returned from the `nls()` function (i.e., by using the S3 `summary()` method). To extract estimated model parameters, use S3 `coef()` method.

To create a simple plot of the model, use S3 `plot()` method, which returns **ggplot2** [@R-ggplot2] object:

@@

Once we have estimated MSS and TAU, we can use `predict_XXX()` family of functions to predict various relationships (i.e., time at distance, acceleration at distance, velocity at time, etc.):

```{r}
# Predict time at distance
predict_time_at_distance(
  distance = split_distance,
  MSS = m1$parameters$MSS,
  TAU = m1$parameters$TAU
)
```

### Air resistance and the calculation of force and mechanical power

To estimate force production at distance or time (using `predict_force_at_distance()` and `predict_force_at_time()` functions), as well as power production (using `predict_power_at_distance()` and `predict_power_at_time()` functions), one needs to take into account the air resistance. Air resistance (N) is estimated using `get_air_resistance()` function, which takes velocity, body mass (kg), body height (m), barometric pressure (Torr), air temperature ($C^\circ$), and wind velocity ($ms^-1$) as parameters (please refer to @arsacModelingEnergetics100m2002, @samozinoSimpleMethodMeasuring2016, and @vaningenschenauCanCyclePower1991 for more information):

@@

```{r}
# To calculate horizontal force produced
predict_force_at_distance(
  distance = split_distance,
  MSS = m1$parameters$MSS,
  TAU = m1$parameters$TAU,
  # Additional parameters forwarded to get_air_resistance
  # Otherwise, defaults are used
  bodymass = 80,
  bodyheight = 1.85,
  barometric_pressure = 780,
  air_temperature = 20,
  wind_velocity = 0.5
)
```

@@

```{r}
df <- predict_kinematics(
  m1,
  max_time = 6,
  frequency = 100,
  # Additional parameters forwarded to get_air_resistance
  # Otherwise, defaults are used
  bodymass = 80,
  bodyheight = 1.85,
  barometric_pressure = 780,
  air_temperature = 20,
  wind_velocity = 0.5
)
```

Plotting the model predictions can be done once we convert data from wide to long with the help of **ggplot2** [@R-ggplot2], **dplyr** [@R-dplyr], **tidyr** [@R-tidyr], and **tidyverse** [@R-tidyverse] packages:

```{r}
require(tidyverse)

variable_names <- colnames(df)

@@

### Utility functions

Another valuable addition for sport scientists and coaches is the ability to determine the distances and times where 90% of maximum sprinting speed is reached, or where peak power is within 90% range. To identify these values, **shorts** package comes with `find_XXX()` family of functions:

```{r}
# Finds distance where 90% of maximum sprinting speed is reached
find_velocity_critical_distance(
  MSS = m1$parameters$MSS,
  TAU = m1$parameters$TAU,
  percent = 0.9
)

# Finds maximal power and distance (this time using air resistance)
find_max_power_distance(
  MSS = m1$parameters$MSS,
  TAU = m1$parameters$TAU,
  # Additional parameters forwarded to get_air_resistance
  # Otherwise, defaults are used
  bodymass = 80,
  bodyheight = 1.85,
  barometric_pressure = 780,
  air_temperature = 20,
  wind_velocity = 0.5
)

# Finds distance over 90% power range
find_power_critical_distance(
  MSS = m1$parameters$MSS,
  TAU = m1$parameters$TAU,
  # Additional parameters forwarded to get_air_resistance
  # Otherwise, defaults are used
  bodymass = 80,
  bodyheight = 1.85,
  barometric_pressure = 780,
  air_temperature = 20,
  wind_velocity = 0.5
)
```

### Mixed-effects model

Sprint performance is often evaluated with a group of athletes (e.g., soccer club) representing a single strata of interest. Sports scientists can estimate individual profiles, or utilize mixed-effects models. To perform mixed-effects models in **shorts** for split times, one can use `mixed_model_using_splits()` function. To demonstrate this functionality, we load the `split_times` dataset provided in the **shorts** package:

```{r}
data(split_times)

# Mixed model
m2 <- mixed_model_using_splits(
  data = split_times,
  distance = "distance",
  time = "time",
  athlete = "athlete",

  # Select random effects
  # Default is MSS and TAU
  random = MSS + TAU ~ 1
)

Additional information about mixed-effects model performed using the `nlme` package [@R-nlme] can be obtained using `summary()` function. S3 method `coef()` when applied on mixed-model result will return both the fixed and random effects.

```{r include=FALSE}
summary(m2)
coef(m2)
```

To create a simple plot of the model, use S3 `plot()` method.

```{r}
plot(m2) + theme_bw(8)
```

The following figure contains kinematics for all athletes in `split_times` dataset. Please note that power calculation takes default parameters for each individual:

```{r}
df <- predict_kinematics(m2, max_time = 10)

variable_names <- colnames(df)

df <- pivot_longer(df, cols = c(-1, -3)) %>%
  mutate(name = factor(name, levels = variable_names))

ggplot(
  filter(df, distance < 40),
  aes(x = distance, y = value, group = athlete, color = athlete)
) +
  theme_bw(8) +
  facet_wrap(~name, scales = "free_y") +
  geom_line(alpha = 0.7) +
  ylab(NULL) +
  xlab("Distance (m)") +
  theme(
    legend.position = "top",
    legend.title = element_blank()
  )
```

## Estimating short sprint parameters using radar gun

Estimation of the short sprint profile using radar gun data takes time as predictor and velocity as the outcome variable. Thus equation \@ref(eq:velocity-time) is used to estimate MSS and TAU.

Let's consider the same example of an athlete with MSS equal to 9 $ms^{-1}$, TAU equal to 1.3, and MAC equal to 6.92 $ms^{-2}$ performing 40m sprint with velocity estimated using radar run (in this case with 1 Hz sampling rate).

```{r}
sprint_time <- seq(0, 6, 1)

sprint_velocity <- c(0.00, 4.83, 7.07, 8.10, 8.59, 8.81, 8.91)

m3 <- model_using_radar(
  velocity = sprint_velocity,
  time = sprint_time
)
```

Both split and radar gun models allow the use of *weighted* non-linear regression. Weighting in regression is utilized when the observations have unequal error variance [@gelmanRegressionOtherStories2020], which can happen due to instrumental error (e.g., a multiplicative error instead of additive error) or due to biological variability (e.g., higher split time variance on shorter distances versus longer distances). In this case, observations with higher error variance get lower weight when fitting the model [@gelmanRegressionOtherStories2020]. According to @gelmanRegressionOtherStories2020, unequal variances are not typically a major issue for the goal of estimating regression parameters, but they can become more important when making predictions about individual observations. Further exploring the topic of unequal error variances is beyond the scope of this paper, but the **shorts** package provides a weighting feature for the potential research of this topic in the future.

Weighted non-linear regression is performed by setting `weights` parameter. For example, we can give more weight to shorter distances or faster velocities:

```{r}
m3_weighted <- model_using_radar(
  velocity = sprint_velocity,
  time = sprint_time,
  weights = 1 / (sprint_velocity + 1)
)
```

### Mixed-effects model

Mixed-effects model using radar data is done using `mixed_model_using_radar()` function. To perform mixed model, let's load data that comes with **shorts** package.

```{r}
data("radar_gun_data")

m4 <- mixed_model_using_radar(
  radar_gun_data,
  time = "time",
  velocity = "velocity",
  athlete = "athlete"
)
```

@@

To create *Force-Velocity Profile* (FVP) using single athlete estimated sprint model parameters (i.e., TAU and MSS), you can use `get_FV_profile()` function. When estimating FVP, athlete body mass (kg) can be set using `bodymass` parameter, while the air resistance parameters can be set using `"..."`, which are forwarded to the `get_air_resistance()` function. Details of the FVP method implemented in the **shorts** package, as well as the interpretation from a sprint training perspective, are covered elsewhere [@haugenPowerForceVelocityProfilingSprinting2020; @morinInterpretingPowerForceVelocityProfiles2016; @morinSimpleMethodComputing2019; @samozinoSimpleMethodMeasuring2016].

```{r}
# To create Force-Velocity Profile
fvp <- get_FV_profile(
  MSS = m1$parameters$MSS,
  TAU = m1$parameters$TAU,
  bodymass = 80,
  # Additional parameters forwarded to get_air_resistance
  # Otherwise, defaults are used
  bodyheight = 1.85,
  barometric_pressure = 780,
  air_temperature = 20,
  wind_velocity = 0.5
)

@@

To plot FVP kinematics and kinetics (which are exactly the same as generated by the `predict_kinematics()` function), use S3 `plot()` function. By default, FVP estimated kinetics are plotted against velocity (on x-axis).

```{r}
plot(fvp) + theme_bw(8)
```

@@

There is a challenge when collecting sprint data that could have a substantial impact on modeled outcomes. To ensure accurate parameter outcomes, the initial force production must be synced with start time [@haugenPowerForceVelocityProfilingSprinting2020; @haugenSprintMechanicalVariables2019]. Below we describe this challenge when using radar guns or timing gates and suggest potential solutions within the **shorts** package.

## Problems with time sync with radar gun

One source of error in the modeled estimation using a radar gun is the time synchronization. In theory, synchronization is ideal when a sprint is initiated at $t=0$ (i.e., $v(t=0) = 0$). In practice, this is often not the case. Let's use our athlete and add and deduct 0.5 s to simulate an error in synchronization and its effect on estimated MSS and TAU.

>>>>

---
title: "**{shorts}**: An R Package for Modeling Short Sprints"
date: "`r Sys.Date()`"
author: 
  - Mladen Jovanović^[Faculty of Sport and Physical Education, University of Belgrade, Serbia, coach.mladen.jovanovic@gmail.com]
  - Jason D. Vescovi^[Faculty of Kinesiology and Physical Education, Graduate School of Exercise Science, Toronto, ON Canada, vescovij@gmail.com]
abstract: >
  Short sprint performance is one of the sport's most distinguishable and admired physical traits. Short sprints have been modeled using the mono-exponential equation that involves two parameters: (1) maximum sprinting speed (MSS) and (2) relative acceleration (TAU). The most common methods to assess short sprint performance are radar guns or timing gates. In this paper, we: 1) provide the **{shorts}** package that can model sprint timing data from these two sources; 2) discuss potential issues with assessing sprint time (synchronization and flying start, respectively); and 3) provide model definitions within the **{shorts}** package to help alleviate errors within the subsequent parameter outcomes.
citation_package: natbib
bibliography: [references.bib, packages.bib]
output:
  bookdown::word_document2:
    number_sections: FALSE
link-citations: yes
editor_options:
  chunk_output_type: console
---

@@

def.chunk.hook <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  ifelse(options$size != "normalsize", paste0("\n \\", options$size, "\n\n", x, "\n\n \\normalsize"), x)
})

@@

# automatically create a bib database for R packages
# knitr::write_bib(c(
#  .packages(), "bookdown", "knitr", "rmarkdown", "shorts", "LambertW", "kableExtra", #"minpack.lm"
# ), "packages.bib")

@@

options(
  digits = 3,
  "width" = 65,
  scipen = 999,
  knitr.kable.NA = ""
)
```

@@

Short sprint performance is one of the sport's most distinguishable and admired physical traits. Short sprints, commonly performed in most team sports (e.g., soccer, field hockey, handball, football, etc.), are defined as maximal running from a stand-still position over a distance that does not result in deceleration at the end. Peak anaerobic power is achieved within the first few seconds (<5 $s$) of maximal efforts [@mangineSpeedForcePower2014]. In contrast, achieving maximal sprint speed varies based on the athlete and sport. For example, track and field sprinters are trained to achieve maximal speed later in a race (i.e., 50-60 $m$) [@ward-smithEnergyConversionStrategies2001], but team sport athletes have sport-specific attributes and reach it much sooner (i.e., 30-40 $m$)[@brownAssessmentLinearSprinting2004]. Regardless of the differences in kinematics between athletes, evaluating short sprint performance is routinely included within a battery of fitness tests for a wide range of sports

The use of force plates is considered the gold standard for assessing the mechanical properties of sprinting; however, there are logistical and financial challenges to capturing the profile of an entire sprint [@morinSimpleMethodComputing2019; @samozinoSimpleMethodMeasuring2016]. Radar and laser technology are frequently used laboratory-grade methods [@buchheitMechanicalDeterminantsAcceleration2014; @edwardsSprintAccelerationCharacteristics2020; @jimenez-reyesRelationshipVerticalHorizontal2018; @marcote-pequenoAssociationForceVelocity2019] but are not ordinarily accessible to practitioners working in sports. Undoubtedly, the most common method available and used to evaluate sprint performance is the use of timing gates. Often multiple gates are positioned at varying distances to capture split times (e.g., 5, 10, 20 $m$), which can now be incorporated into the method for determining sprint mechanical properties [@morinSimpleMethodComputing2019; @samozinoSimpleMethodMeasuring2016]. This approach presents an advantage to practitioners using the outcomes to describe individual differences, quantify the effects of training interventions, and better understand performance limiting factors.

The **{shorts}** package [@R-shorts], written in the R language [@R-base], represents an open-source tool to help sports scientists translate raw timing data into detailed mechanical outcomes through mathematical modeling [@morinSimpleMethodComputing2019; @samozinoSimpleMethodMeasuring2016]. To the best of our knowledge, scientist, researchers, and coaches have been performing short sprints modeling using the built-in solver function of Microsoft Excel (Microsoft Corporation, Redmond, Washington, United States) [@morinSpreadsheetSprintAcceleration2017; @morinSpreadsheetSprintAcceleration2019; @stenrothForcevelocityProfilingIce2020; @stenrothSpreadsheetSprintAcceleration2020; @samozinoSimpleMethodMeasuring2016; @clarkNFLCombine40Yard2017; @morinSimpleMethodComputing2019], which makes the **{shorts}** package a major improvement in ease-of-use, speed, transparency, reproducibility, and more feature-rich model fitting.

The current paper will explain one commonly used mathematical equation to model short sprints, modeling applications using the **{shorts}** package, issues arising during measurement and estimation, and potential solutions to those problems.

@@

Short sprints have been modeled using the mono-exponential equation \@ref(eq:velocity-time), originally proposed by @doi:10.1098/rspb.1927.0035 and more recently popularized by @clarkNFLCombine40Yard2017 and @samozinoSimpleMethodMeasuring2016. Equation \@ref(eq:velocity-time) represents the function for instantaneous horizontal velocity $v$ given the time $t$ and two model parameters:

@@

The parameters of equation \@ref(eq:velocity-time) are *maximum sprinting speed* (MSS; expressed in $ms^{-1}$) and *relative acceleration* (TAU; expressed in $s$). Mathematically, TAU represents the ratio of MSS to initial acceleration (MAC; *maximal acceleration*, expressed in $ms^{-2}$) (equation \@ref(eq:maximal-acceleration)). Given the equation \@ref(eq:velocity-time), TAU can be interpreted as the time required to reach a sprinting velocity equal to `r round(100 * (1 - exp(-1)), 1)`% of MSS.

@@

Although TAU is used in the equations and later estimated, we prefer MAC since it is easier to grasp, particularly for less math-inclined coaches.

@@

Let us consider four athletes with different levels of MSS (high versus low maximal sprinting speed) and MAC (high versus low maximal acceleration; as mentioned previously, using MAC is preferred over using TAU):

@@

knitr::kable(
  athletes,
  #caption = "Four athletes with different MSS and MAC parameters.",
  digits = 2,
  booktabs = TRUE
)
```

Next figure depicts distance, velocity, and acceleration over time (from 0 to 6 $s$):

```{r four-athletes-kinematics, echo=FALSE}
kinematics <- expand_grid(
  athletes,
  time = seq(0, 6, length.out = 1000)
) %>%
  mutate(
    distance = predict_distance_at_time(time, MSS, MAC),
    velocity = predict_velocity_at_time(time, MSS, MAC),
    acceleration = predict_acceleration_at_time(time, MSS, MAC)
  )

@@

Plotting acceleration against velocity in the next figure, we will get the *Acceleration-Velocity Profile*, which is linear, according to the mathematical model. If the athlete's body mass (kg) is known, as well as additional air resistance parameters (see [Air resistance and the calculation of force and mechanical power] section of this paper), *Force-Velocity Profile* can be estimated (see [Force-Velocity profile] section of this paper).

```{r four-athletes-profile, echo=FALSE}
gg <- ggplot(
  kinematics,
  aes(x = velocity, y = acceleration, color = Athlete)
) +
  theme_bw(8) +
  geom_line(alpha = 0.7) +
  ylab(expression("Acceleration (ms"^-2 * ")")) +
  xlab(expression("Velocity (ms"^-1 * ")")) +
  xlim(c(0, 12)) +
  theme(
    legend.position = "top",
    legend.title = element_blank()
  )

@@

# Estimation using **{shorts}** package

Short sprints profiling is usually performed by: (1) measuring split times using timing gates (i.e., positioned at various distances, e.g., 5, 10, 20, 30, 40 $m$), or (2) getting a velocity trace using a radar gun. Estimation of MSS and TAU parameters from equation \@ref(eq:velocity-time) is performed in **{shorts}** package for the **R language** [@R-base] using *non-linear least squares regression* implemented in the `nlsLM()` function from the **{minpack.lm}** package [@R-minpack.lm].

@@

For timing gates split times, distance is a predictor, and time is the outcome variable. Thus, equation \@ref(eq:distance-time) becomes:

@@

In equation \@ref(eq:time-distance), $W$ represents Lambert's W function [@R-LambertW]. Researchers often incorrectly use the equation \@ref(eq:distance-time) [@morinSpreadsheetSprintAcceleration2017; @morinSpreadsheetSprintAcceleration2019; @stenrothSpreadsheetSprintAcceleration2020], in which the time is the predictor and distance is the outcome variable, instead of the statistically correct equation \@ref(eq:time-distance). This practice should be avoided because switching the predictor and outcome variables in the regression model might produce biased estimated parameters [@motulskyIntuitiveBiostatisticsNonmathematical2018, p. 341]. These biases might not necessarily be significant with the short sprints profiling, but it is nevertheless a bad statistical practice.

Here is an example using a built-in dataset. We are going to filter out one athlete (e.g., John) with the help of the **tidyverse** [@R-tidyverse] package and estimate MSS, TAU, and MAC parameters using the `model_timing_gates()` function:

```{r}
require(shorts)
require(tidyverse)

# Load built-in dataset
data(split_times)

# Filter timing gates splits for John
john_TG_data <- split_times %>%
  filter(athlete == "John")

john_TG_data

# Estimate John's MSS, TAU, MAC, and PMAX
m1 <- model_timing_gates(
  distance = john_TG_data$distance,
  time = john_TG_data$time
)

@@

Maximal relative power (PMAX) from the output is estimated using $\frac{MSS \times MAC}{4}$, disregarding the air resistance.

Besides providing *residual standard error* (RSE), **{shorts}** functions provide additional model fit estimators. Additional information can be gained by exploring the returned object, particularly the object returned from the `nlsLM()` function (i.e., by using the S3 `summary()` method). To extract estimated model parameters, use S3 `coef()` method.

To create a simple plot of the model, use the S3 `plot()` method, which returns **ggplot2** [@R-ggplot2] object:

@@

Once we have estimated MSS and MAC, we can use `predict_XXX()` family of functions to predict various relationships (i.e., time at distance, acceleration at distance, velocity at time, etc.):

```{r}
# Predict time at distance
predict_time_at_distance(
  distance = john_TG_data$distance,
  MSS = m1$parameters$MSS,
  MAC = m1$parameters$MAC
)
```

## Estimating short sprint parameters using the radar gun

Estimating the short sprint profile using radar gun data takes time as a predictor and velocity as the outcome variable. Estimation using radar gun data is implemented in the **{shorts}** package using `model_radar_gun()` function.

Here is an example using built-in dataset:

```{r}
# Load built-in dataset
data(radar_gun_data)

# Filter radar gun data for John
john_RG_data <- radar_gun_data %>%
  filter(athlete == "John")

head(john_RG_data)

# Estimate John's MSS, TAU, MAC, and PMAX
m2 <- model_radar_gun(
  velocity = john_RG_data$velocity,
  time = john_RG_data$time
)

There is the additional parameter in the output, TC, that is estimated using the `model_radar_gun()` function. The function of this parameter is explained in the [Problems with time sync with the radar gun] section of this paper.

Both timing gates and radar gun models allow *weighted* non-linear regression. Weighting in regression is utilized when the observations have unequal error variance [@gelmanRegressionOtherStories2020], which can happen due to instrumental error (e.g., a multiplicative error instead of additive error) or due to biological variability (e.g., higher split time variance on shorter distances versus longer distances). In this case, observations with higher error variance get lower weight when fitting the model [@gelmanRegressionOtherStories2020]. According to @gelmanRegressionOtherStories2020, unequal variances are not typically a significant issue for the goal of estimating regression parameters, but they can become more important when making predictions about individual observations. Further exploring the topic of unequal error variances is beyond the scope of this paper. However, the **{shorts}** package provides a weighting feature for potential research of this topic in the future.

Weighted non-linear regression is performed by setting the `weights` parameter. For example, we can give more weight to shorter distances or faster velocities:

```{r}
m2_weighted <- model_radar_gun(
  velocity = john_RG_data$velocity,
  time = john_RG_data$time,
  weights = 1 / (john_RG_data$velocity + 1)
)

m2_weighted
```

## Air resistance and the calculation of force and mechanical power

To estimate force production at distance or time (using `predict_force_at_distance()` and `predict_force_at_time()` functions), as well as power production (using `predict_power_at_distance()` and `predict_power_at_time()` functions), one needs to take into account the air resistance. Air resistance (measured in Newtons, $N$) is estimated using `get_air_resistance()` function, which takes velocity, body mass ($kg$), body height ($m$), barometric pressure ($Torr$), air temperature ($C^\circ$), and wind velocity ($ms^-1$) as parameters (please refer to @arsacModelingEnergetics100m2002, @samozinoSimpleMethodMeasuring2016, and @vaningenschenauCanCyclePower1991 for more information):

@@

```{r}
# To calculate horizontal force produced
predict_force_at_distance(
  distance = john_TG_data$distance,
  MSS = m1$parameters$MSS,
  MAC = m1$parameters$MAC,
  # Additional parameters forwarded to get_air_resistance
  # Otherwise, defaults are used
  bodymass = john_TG_data$bodyweight,
  bodyheight = 1.85,
  barometric_pressure = 780,
  air_temperature = 20,
  wind_velocity = 0.5
)
```

@@

```{r}
df <- predict_kinematics(
  m1,
  max_time = 6,
  frequency = 100,
  # Additional parameters forwarded to get_air_resistance
  # Otherwise, defaults are used
  bodymass = john_TG_data$bodyweight[1],
  bodyheight = 1.85,
  barometric_pressure = 780,
  air_temperature = 20,
  wind_velocity = 0.5
)
```

Plotting the model predictions can be done once we convert data from wide to long with the help of **ggplot2** [@R-ggplot2], **dplyr** [@R-dplyr], and **tidyr** [@R-tidyr] packages (loaded already using the **tidyverse** [@R-tidyverse] package):

```{r}
variable_names <- colnames(df)

@@

## Utility functions

Another valuable addition for sports scientists and coaches is the ability to determine the distances and times where 90% of maximum sprinting speed is reached, or where peak power is within 90% range. To identify these values, the **{shorts}** package comes with the `find_XXX()` family of functions:

```{r}
# Finds distance where 90% of maximum sprinting speed is reached
find_velocity_critical_distance(
  MSS = m1$parameters$MSS,
  MAC = m1$parameters$MAC,
  percent = 0.9
)

# Finds maximal power and distance (this time using air resistance)
find_max_power_distance(
  MSS = m1$parameters$MSS,
  MAC = m1$parameters$MAC,
  # Additional parameters forwarded to get_air_resistance
  # Otherwise, defaults are used
  bodymass = 80,
  bodyheight = 1.85,
  barometric_pressure = 780,
  air_temperature = 20,
  wind_velocity = 0.5
)

# Finds distance over 90% power range
find_power_critical_distance(
  MSS = m1$parameters$MSS,
  MAC = m1$parameters$MAC,
  # Additional parameters forwarded to get_air_resistance
  # Otherwise, defaults are used
  bodymass = 80,
  bodyheight = 1.85,
  barometric_pressure = 780,
  air_temperature = 20,
  wind_velocity = 0.5
)
```

@@

To create *Force-Velocity Profile* (FVP) using single athlete estimated sprint model parameters (i.e., MSS and MAC), you can use the `make_FV_profile()` function. When estimating FVP, athlete body mass (kg) can be set using `bodymass` parameter, while the air resistance parameters can be set using `"..."`, which are forwarded to the `get_air_resistance()` function. Details of the FVP method implemented in the **{shorts}** package and the interpretation from a sprint training perspective are covered elsewhere [@haugenPowerForceVelocityProfilingSprinting2020; @morinInterpretingPowerForceVelocityProfiles2016; @morinSimpleMethodComputing2019; @samozinoSimpleMethodMeasuring2016].

```{r}
# To create Force-Velocity Profile
fvp <- make_FV_profile(
  MSS = m1$parameters$MSS,
  MAC = m1$parameters$MAC,
  bodymass = 80,
  # Additional parameters forwarded to get_air_resistance
  # Otherwise, defaults are used
  bodyheight = 1.85,
  barometric_pressure = 780,
  air_temperature = 20,
  wind_velocity = 0.5
)

@@

To plot FVP kinematics and kinetics (which are the same as generated by the `predict_kinematics()` function), use the S3 `plot()` function. FVP estimated kinetics are default plotted against velocity (on the x-axis).

```{r results='hide', fig.show='hide'}
plot(fvp) + theme_bw(8)
```

@@

There is a challenge when collecting sprint data that could substantially impact modeled outcomes. To ensure accurate parameter outcomes, the initial force production must be synced with the start time [@haugenPowerForceVelocityProfilingSprinting2020; @haugenSprintMechanicalVariables2019]. Below we describe this challenge when using radar guns or timing gates and suggest potential solutions within the **{shorts}** package.

## Problems with time sync with the radar gun

Time synchronization is one error in the modeled estimation using a radar gun. In theory, synchronization is ideal when a sprint is initiated at $t=0$ $s$ (i.e., $v(t=0) = 0$ $ms^-1$). In practice, this is often not the case. This might happen when the measurement starts before the sprint starts. Although these data should be *trimmed*, the time variable might not be synchronized perfectly (i.e., sprint starts exactly at $t=0$ $s$) with the sprint initiation afterward.

The potential solution incorporated into the **{shorts}** package involves estimating the *time correction* (TC) parameter using equation \@ref(eq:velocity-time-correction). TC parameter serves as *intercept*, similar to linear regression, and allows the model to be fitted correctly in the scenario explained.
